Ordering algorithm used in ExpressionView

### 2010-03-14 RExpressionView/inst/doc/ExpressionView.ordering.Rnw
ExpressionView:  Removed old images.

The \code{C++} code contains the functions \code{reposition} and \code{exchange} that 
execute these two basic moves.

Before optimizing the arrangement, the matrix is simplified (\code{simplify}) to get rid of duplicate slots.
The multiplicity of each slot is taken into account when calculating the score $Q$. 
At the end of the ordering the initial dimensions of the data are restored with the \code{complexify} 
function.

@@

\bibliographystyle{unsrt}
\bibliography{ExpressionView}

>>>>

Internally, the data is represented as a binary matrix (i.e., elements are either 0 or 1) with dimensions $m$ $\times$ $n$, where the number of rows $m$ correspond to the number of genes or conditions and the number of columns $n$ is determined by the number of clusters.

The ordering is done in four steps:
\begin{enumerate}
	\item{\code{simplify}: Reduces the dimensions of the problem by excluding duplicate rows. The multiplicity of each slot is taken into account when calculating the score $Q$.}
	\item{\code{prearrange}: Tries to find a good starting point for the optimization by creating a new order, starting from the first row and adding the consecutive rows one after the other, placing them at the optimal position. For smaller samples, the fitness function is used, while for larger samples only the similarity (scalar product) between rows is taken into account.}
	\item{\code{arrange}: Uses the brute-force approach to further increase the alignment score.}
	\item{\code{complexify}: Reinstates the original dimensions of the problem by inserting the omitted rows at the newly found position.}
\end{enumerate}

If the user specifies a time limit, the algorithm stops the alignment after the elapsed interval and returns the best alignment. Otherwise, it continues until it can no longer improve the alignment score.

The ordering iterates over the clusters and first calls the function \code{reposition}, taking \code{cluster} as argument. It goes through all the rows contained in a given cluster and tries to shift them to a better position. A contiguous group of rows can be repositioned at once or split up in subgroups. Empirically, it turned out that the alignment is fastest by starting the repositioning with a single row, increasing the length of the group only if no better position could be found. The opposite approach can be used by replacing the \emph{start forward} section on line 545 by the \emph{start backwards} code on line 556.

After repositioning, the main loop calls the exchange function, taking \code{cluster} and \code{what} as argument. It exchanges rows contained (\code{what = 1}) / not contained (\code{what = 0}) in a given cluster. It randomly picks \code{maxslots} number of rows (currently six), generates all possible permutations, calculates the corresponding alignment scores, and retains the best order. This realignment is repeated \code{maxiter} times, so that on average every slot is part of one optimization.

The loop over clusters calls the \code{reposition} function first for all the rows not contained in a given cluster and afterwards for the remaining ones. The main loop is repeated until the maximal computation time is reached or no improved alignment could be found anymore.

@@

%\bibliographystyle{unsrt}
%\bibliography{ExpressionView}
